Basic concepts and tools for the Toki Pona minimalist and constructed language

### 2018-07-01 article/acm/samplebody-journals.tex
arrumacao inicial do repositorio com formatacao p acm do artigo

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\usepackage{tipa}
\newcommand{\ttt}[1] {
	\texttt{<#1>}}
\newcommand{\tttt}[1]{\texttt{#1}}
% \usepackage{fontspec}
% \usepackage[mathletters]{ucs}
% \newcommand{\ttt}[1] {
% 	\texttt{<#1>}}
% \newcommand{\tttt}[1]{\texttt{#1}}
\usepackage{subcaption}

\begin{document}
\title{Basic concepts and tools for the Toki Pona minimalist and
constructed language: description of the language and main issues;
Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
% Anthropological computation
\author{Renato Fabbri, \texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
A minimalist constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimalist both in the vocabulary
(with only 14 letters and 124 words)
and in the $\approx10$ syntax rules.
The language is useful for being a used and somewhat established
minimalist conlang with at least hundreds of fluent speakers.
In this article, we describe current concepts and resources
for TP,
and make available Python scripted routines for
the analysis of the language,
the synthesis of texts, the specification of syntax highlighting
schemes, and the achievement of a preliminary TP
Wordnet~\cite{wordnet}.
We focus on the analysis of the basic vocabulary,
as corpus analyses were found~\cite{corpus}.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor~\cite{vim}.
The tentative TP Wordnet is made available in three forms that
reflect the choices of the synsets related to each word.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing and syntax highlighting the
TP language.
  % subj (li), obj (e), prepositions, qualification (word concatenation), 
  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
\end{abstract}
{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona

@@

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3, coordinated by Prof. Dr. Maria Cristina Ferreira de Oliveira);
Toki Pona, Python and Vim authors, developers and literature maintainers;
Toki Pona user and speaker communities. 
I thank Mario Alzate for introducing me to Toki Pona,
and Silverio Guazzelli Donatti for the stimulus that resulted
from the usage of Toki Pona for chatting.

@@

\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem{wordnet}
  Fellbaum, C. (1998). WordNet. John Wiley \& Sons, Inc..
\bibitem{corpus}
    Henry, J. lipu pi toki pona pi jan Jakopo. Available at         
    \url{http://jimhenry.conlang.org/conlang/tokipona/tokipona.htm} 
\bibitem{vim}
    Fabbri, R. An anthropological account of the Vim editor:
features and tweaks after 10 years of usage. Available at         
    \url{https://github.com/ttm/vim/raw/master/article/article.pdf} 
\bibitem{sapWho}
  O'Neill, S. P. (2015). Sapir–Whorf Hypothesis. The International
    Encyclopedia of Language and Social Interaction.
\bibitem{tpLang}
	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
\bibitem{kama}
  Knight, B. (2017) o kama sona e toki pona! Available at:
    \url{http://tokipona.net/tp/janpije/okamasona.php}
\bibitem{tp3}
  Blahuš, M. (2011). Toki pona–eine minimalistische Plansprache.
    Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
\bibitem{tp4}
  Schneider, S. (2017). nasin toki pi toki pona. Available at
  \url{https://github.com/stefichjo/toki-pona}
\bibitem{interview}
  Lang, S. and Broholm, K. AFP 20 – Sonja Lang: Toki Pona, Conlanging
    and the meaning of life (Sonja Lang interview). Available at
    \url{https://actualfluency.com/afp-20-sonja-lang-toki-pona-conlanging-meaning-life/}
\bibitem{memrise}
  Toki Pona Community. (2017). Various Memrise courses on Toki Pona.
    \url{https://www.memrise.com/courses/english/?q=toki+pona}
\bibitem{gdoc}
  Toki Pona Community (2017). toki pona central hub preparation.
    Available at
    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
\bibitem{wikiToki}
  Toki Pona. (2017, December 11). In Wikipedia, The Free Encyclopedia.
    Retrieved 22:48, December 14, 2017, from
    \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
\bibitem{conlanWikip}
  Constructed language. (2017, November 1). In Wikipedia, The Free Encyclopedia. Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
\bibitem{artlanWikip}
  Artificial language. (2017, November 26). In Wikipedia, The Free
  Encyclopedia. Retrieved 23:00, December 14, 2017, from
  \url{https://en.wikipedia.org/w/index.php?title=Artificial_language&oldid=812127746}
\bibitem{akesiWawa}
  Fabbri, R. (2017, November 6). o awen, akesi wawa.
  In the group toki pona taso. Available at
  \url{https://www.facebook.com/groups/tokiponataso/permalink/1895942817391318/}
\bibitem{janKipo}
  Clifford, J. et al. (2017, December 15). Considerations about Toki Pona
  phrases.
  In the Facebook group 'toki pona'. Available at
  \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
\bibitem{tokipona}
	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
\bibitem{tpNetSH}
  Knight, B. (2017). Online gadget for syntax highlighting Toki Pona
  texts. . Available at
  \url{http://tokipona.net/tp/DisplayText.aspx}
\bibitem{stack}
  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
    Software Engineering in Stack Exchange. Available at
    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
\bibitem{solarized}
  Schoonover, E. Solarized color scheme. Available at
  \url{http://ethanschoonover.com/solarized}
\bibitem{dv1}
	Munzner, T. (2014). Visualization analysis and design. CRC press.
\bibitem{dv2}
  Telea, A. C. (2014). Data visualization: principles and practice.
  CRC Press.
\bibitem{dv3}
  Ward, M. O., Grinstein, G., \& Keim, D. (2010). Interactive data
  visualization: foundations, techniques, and applications. CRC Press.
\bibitem{dv4}
  Ware, C. (2012). Information visualization: perception for design.
  Elsevier.
\bibitem{blue}
  Mehta, R., \& Zhu, R. J. (2009). Blue or red? Exploring the effect of
  color on cognitive task performances. Science, 323(5918), 1226-1229.
\bibitem{blue2}
  Viola, A. U., James, L. M., Schlangen, L. J., \& Dijk, D. J. (2008).
  Blue-enriched white light in the workplace improves self-reported
  alertness, performance and sleep quality. Scandinavian journal of
  work, environment \& health, 297-306.
\bibitem{fonts}
  Spolsky, A. J. (2008). User interface design for programmers.
  Apress.
\bibitem{mass}
	Fabbri, R., Vieira, V. , Pessotti, A. C. S., Corrêa, D. C., Oliveira Jr, O. N. O. Musical elements in the discrete-time representation of sound (2017). arXiv preprint arXiv:1412.6853. Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
\bibitem{wikiPoetry}
Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
    Retrieved 19:46, December 3, 2017, from
    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
\bibitem{tokisona}
  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
  Available at \url{http://tokisona.github.io/}
\bibitem{Wikipesija}
  Toki Pona community (2017). Wikipesija. Available at:
    \url{http://tokipona.wikia.com}
\end{thebibliography}
\end{document}

>>>>

@@

@@

\begin{acks}

The author thanks
FAPESP (project 2017/05838-3, coordinated by Prof. Dr. Maria Cristina Ferreira de Oliveira);
Toki Pona, Python and Vim authors, developers and literature maintainers;
Toki Pona user and speaker communities; 
Mario Alzate for introducing me to Toki Pona,
and Silverio Guazzelli Donatti for the stimulus that resulted
from the usage of Toki Pona for chatting.

\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-bibliography}
